---
title: "Regression Modelling in R"
author: "Chris Mainey chris.mainey@uhb.nhs.uk"
date: "2019/09/02 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: "libs/HED.css"
    lib_dir: libs
    seal: false
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
class: center


```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "center"
)
```



<br><br><br><br>

## Regression Modelling in `R`

<br>

.pull-left[
__Chris Mainey__
<p style="font-size:26;")>Intelligence Analyst<br>
<b>University Hospitals Birmingham 
NHS Foundation Trust</b></p>
<br>
<span style="font-size:26;">[chris.mainey@uhb.nhs.uk](mailto:chris.mainey@uhb.nhs.uk)</span><br>
`r icon::fa("twitter")` <a href="https://twitter.com/chrismainey?s=09" style="line-height:2;">@chrismainey</a>
]

.pull-right[
<img src= "./man/figures/NHSR-logo.png" width=80% height=80%>
<br><br>
<img src= "./man/figures/HI.png" width=80% height=80%>
]


---

# Workshop Overview

- Correlation
- Linear Regression
 - Specifying a model with `lm`
 - Interpreting the model output
 - Assessing model fit
- Multiple Regression
- Prediction
- Generalized Linear Models using `glm`
 - Logistic Regression
 
--

<br>
___Mixture of theory, examples and practical exercises___

---

# Relationships between variables

If two variables are related, we use the described them are 'correlated'.

--

Often interested in strength of association

--

<br><br>
Two techniques commonly used to investigate:

--

+ ___'Correlation'___ show strength of association

--

+ ___'Regression'___ estimates how one variable, $x$ (or more than one), predicts another, $y$ .

--

<br><br>
Sometimes the effects of other variables interact/mask this ("confounding")

---

```{r lmsetup, echo=FALSE,include=FALSE}
set.seed(222)
x <- rnorm(50, 10, 4)
y <- runif(50, min = 0.5, 10) + (1.25*x)
z <- predict(lm(y~x))
set.seed(222)
library(ggplot2)
```

## Example:
```{r lm1, fig.retina= 2, fig.align="center", fig.height=4.6, fig.width=6}
x <- rnorm(50, 10, 4)
y <- runif(50, min = 0.5, 10) + (1.25*x)

a<-ggplot(data.frame(x,y,z), aes(x=x,y=y))+
  geom_point(col="dodgerblue2", alpha=0.65)
a
```

---

# Correlation

- Measured with a correlation coefficient.
 - 'Pearson' is the most common

- Range:
 - -1 Perfect Negative Correlation
 - 0 No Correlation
 - 1 Perfect positive Correlation
<br><br>
 
![](./man/figures/correlation-examples.svg)


.footnote[
Graphic from:
Pierce, Rod,  2018, 'Correlation', Math Is Fun, Available at: <http://www.mathsisfun.com/data/correlation.html>. [Accessed 8 Jul 2019]
]
---

# Correlation in R

Lets check the correlation in our generated data:

```{r correlation}
cor(x, y)

cor.test(x,y)

```

+ `cor.test` is a correlation and a t-test.
+ Different types of correlation coefficient, default is 'Pearson'
+ Doesn't work for different distributions, data types or more variables

---

### Regression models (1)

Regression gives us more options than correlation:

```{r lm3, fig.retina=2, fig.align="center", fig.height=4.8, fig.width=6}
z <- lm(y~x)

print(a<- a + geom_smooth(method="lm", col="red", linetype=2))
```


$$y= \alpha + \beta x + \epsilon$$

---

### Ordinary Least Squares 'OLS'

+ Uses the residual error (ε) to work out the position of the line.
--

+ 'Residual'; distance between prediction and data point.
--

+ Sum would be zero, square and minimise 'sum of the squares'
--

```{r lm2, fig.retina=2, fig.align="center", fig.height=5, fig.width=6}
a + geom_segment(aes(xend=c(x), yend=c(z)), size=0.5, arrow=arrow(length=unit(0.2,"cm")))
```

---

## Regression models (3)

+ We created an `lm` object called `z`.  You can then use that object with other functions like `summary`, `predict`, `plot` etc.


```{r sum_reg}
summary(z)
```


---

## Regression models (4)

+ Usually view model details with `summary`, but I'm using a function from `sjPlot` here.
```{r lm4}
sjPlot::tab_model(z, show.df = TRUE, show.obs = TRUE)

```
--

<br>
+ We can test fit using f-tests, prediction error or the R<sup>2</sup> .
+ R<sup>2</sup> is the proportion of variation in $y$, explained by $x$.

---

## Regression diagnostics (5)

- A common check is to plot residuals:
```{r lmplot, fig.retina=2, fig.align="center", fig.height=5.5, fig.width=8, fig.keep = 'last'}
par(mfrow=c(2,2))
plot(z)
```


---
class: center, middle

# Exercise 1:  Correlation and Linear regression

---

## What about non-linear data?

- Our data are not necessarily linear. Death is binary, LOS is a count etc.

- We can extend linear models idea for this as a Generalized Linear Model (GLM)

$$\large{g(\mu)= \alpha + \beta x}$$
Where $\mu$ is the expectation of Y, and $g$ is the link function

- The link function transforms the data before fitting a model

- Can't use OLS for this, so we use 'maximum-likelihood'

- Many of the methods for `lm` are common to `glm`, but we can't use R<sup>2</sup>

- Other measures include AUC ('C-statistic'), and AIC or likelihood ratio tests.

---

## Generalized Linear Models

- Use a `family` argument with `poisson` for counts or `binomial` for binary outcomes.
- Plotting residuals is not straight forward, and less common with `glm`

```{r glm1, collapse=TRUE}
library(NHSRdatasets)
data(LOS_model)
glm_binomial <- glm(Death ~ Age + LOS, data=LOS_model, family="binomial")

sjPlot::tab_model(glm_binomial, show.df = TRUE, show.obs = TRUE, show.se = TRUE, show.r2 = FALSE)

ModelMetrics::auc(glm_binomial)

```

---

## Interactions

- 'Interactions' are where predictor variables affect each other.
--

- HSMR has an interaction between co-morbidity and age. So co-morbidities have different effects related to age.
--

- Can add these with `*` or `:` (check help for which to use)
--

```{r glm2}
glm_binomial2<- glm(Death ~ Age + LOS + Age*LOS, data=LOS_model, family="binomial")

sjPlot::tab_model(glm_binomial2, show.df = TRUE, show.obs = TRUE, show.se = TRUE, show.r2 = FALSE)

ModelMetrics::auc(glm_binomial2)

```

---

# Interpretation

+ Our model coefficients in `lm` were straight-forward multipliers
+ `glm` is similar, but it is on the scale of the link-function.
 + log scale for `poisson` models, or logit (log odds) scale for `binomial`
+ Common to transform output back to response scale
+ giving Incident Rate Ratios for `poisson`, or Odds Ratios for `binomial`

```{r coeftransform, warning=FALSE, message=FALSE}
cbind(Link=coef(glm_binomial2), Response=exp(coef(glm_binomial2)))

```

---

## Prediction (1)

- We can then use our model to predict our expected Y:
- Need to decide what scale to predict on: `link` or `response`

```{r glm3}
library(dplyr)
LOS_model$preds <- predict(glm_binomial2, type="response")

top_n(LOS_model,5) %>% knitr::kable(format = "html")
```

---

## Prediction (2)

- Lets see the 10 cases with the highest predicted risk of death:

```{r glm4}
LOS_model %>% arrange(desc(preds)) %>% top_n(10) %>% knitr::kable(format = "html")

```

---
class: middle

# Exercise 2: Building a GLM

---
# Summary

- Correlation shows the direction and strength of association
- Regression allows us to quantify the relationships
- We can use a single, or multiple, predictors 
- Regression coefficents explain how much a change in $x$ affects $y$
- R<sup>2</sup> is a common measure of in linear models, C-statistic/AUC/ROC in logistic moels
- Prediction error is another useful metric
- Generalized Linear Model (`glm`) allow linear models on a tranfomred scale, e.g. logositc regression for binary variables

---
class: middle

# Exercise 3: Examining Frammingham data

---
