---
title: "Regression Commands"
author: "Chris Mainey"
date: "14/10/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  fig.align = "center"
)
```


The tutorial on regression model uses many different command, but most of them are simple and based on default `R` objects and methods.  Never-the-less, it takes time to learn, search and try these commands out, so this document is a summary of the commands show in the slides.  The will use the `LOS_model` dataset from the `NHSRdatasets` package, but will not usually show the output of each command.

You can install/load the package and data with:
```{r pckload, message=FALSE, warning=FALSE}
install.packages("NHSRdatasets")

library(NHSRdatasets)
data(LOS_model)
```

# Correlaion

Although we are not using correlation in the practical, it is mentioned inthe slides.

```{r cars}
# Correlation coefficient for Age and LOS
cor(LOS_model$Age, LOS_model$LOS)

# With a significane test:
cor.test(LOS_model$Age, LOS_model$LOS)

# Correaltion matrix plot of the data. frame (numeric columns 3 - 5)
library(corrplot)

M <-cor(LOS_model[3:5])
corrplot(M, type="upper", order="hclust")

```

## Linear regression

Linear regression is the building block for what we are looking at, and we will treate regressions as 'object', i.e. assinge each regression to a variable.

```{r reg1}}
#Linear model
lm(LOS ~ Age, data=LOS_model)

# Linear model assigned to variable
mod1 <- lm(LOS ~ Age, data=LOS_model)


```

